UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-07-04 ms/main.tex
edits up to line 85

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference}.

Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations. If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ ranging from hundreds of thousands to hundreds of millions. Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.

@@

The amount of information encoded in the lower-variance principal components increases with sample size, so researchers typically examine multiple two-dimensional projections to explore data. Some features may be hidden by the projections or hard to interpret. To display more of the high-dimensional features of the data in a two dimensional figure, we can use non-linear transformations that seek to preserve the local structure of the data. A popular method for visualization is t-distributed stochastic neighbour embedding (t-SNE)\cite{maaten2008visualizing}. t-SNE has been used before to visualize SNPs\cite{platzer2013visualization}; using data from the 1000 Genomes Project (1KGP)\cite{10002015global}, it groups individuals corresponding roughly to their continent of origin, with smaller ethnic sub-groups visible within the larger continental clusters\cite{li2017tsne}. However, t-SNE struggles with very large datasets, when a large number of locally optimal configurations make convergence to a globally satisfying solution difficult.

@@

\subsection*{Results}
\subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}. Figure~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and UMAP with PCA pre-processing. Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters. Using either method on the top principal components leads to distinct population clusters and less defined continental structure. Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}). Patterns are evident even when using UMAP on the last components (figure~\ref{fig:supp_1kgp_3350}).

Focusing on UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries. British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry. Toscani and Iberian individuals form a cluster reflecting their Mediterranean heritage. African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster. Three East Asian clusters appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population. Other clusters are comprised of Colombians and Peruvians, the Esan and Yoruba populations of Nigeria, and several South Asian populations. Within population clusters, family members were projected near each other but did not form disjoint clusters unless UMAP was parameterized to use a minimal number of neighbours (figure~\ref{fig:supp_1kgp_families}).

>>>>

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference} \sgcomment{This is a lone reference for a broad topic... Are there more recent reviews perhaps? Else add a couple of recent references eg by the novembre lab?}.

Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations. If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of copies of the derived alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ ranging from hundreds of thousands to hundreds of millions. Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.

@@

The amount of information encoded in the lower-variance principal components increases with sample size, so researchers typically examine multiple two-dimensional projections to explore data. In this process, finer features of the data may be hidden by the projections or hard to interpret. To display finer features of the data in a two dimensional figure, we can use non-linear transformations that emphasize the local structure of the data. A popular method for such visualization is t-distributed stochastic neighbour embedding (t-SNE)\cite{maaten2008visualizing}. t-SNE has been used before to visualize SNPs\cite{platzer2013visualization}; using data from the 1000 Genomes Project (1KGP)\cite{10002015global}, it groups individuals corresponding roughly to their continent of origin, with smaller ethnic sub-groups visible within the larger continental clusters\cite{li2017tsne}. However, t-SNE struggles with very large datasets, when a large number of locally optimal configurations make convergence to a globally satisfying solution difficult.

@@

\subsection*{Results}
\subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}. Figure~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and UMAP with PCA pre-processing. Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters. Using either method on the top principal components leads to distinct population clusters and less defined continental structure. Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}). To investigate the population information contained in low-variance PCs, we performed UMAP on data projected onto PCs 100 to 3450 (i.e., without information about the leading 99 PCs). Figure~\ref{fig:supp_1kgp_3350} shows that population structure is still clearly visible. \sgcomment{Note figure S4 lacks a color code -- at least point out that it uses the same coloring as X}

Focusing on UMAP with the leading 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries. British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry. Toscani and Iberian individuals form a cluster reflecting their Mediterranean heritage. African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster. Three East Asian clusters appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population. Other clusters are comprised of Colombians and Peruvians, the Esan and Yoruba populations of Nigeria, and several South Asian populations. Within population clusters, family members were projected near each other but did not form disjoint clusters unless UMAP was parameterized to use a minimal number of neighbours (figure~\ref{fig:supp_1kgp_families}).
